Efficient forwards-time simulation with ancestral recombination graphs

### 2017-09-13 forwards_paper.tex
added example tree seq

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath, amssymb}
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{listings}
\bibliographystyle{plainnat}
\usepackage[T1]{fontenc}

@@

However, it is thought that much of that variation is selectively neutral (or nearly so).
By definition, the alleles carried by individuals in a population at neutral sites
do not affect the population process.
For this reason, if one records the entire genealogical history of a population over the course of a simulation,
one can lay down neutral mutations on top of that history afterwards,
without loss of generality.
Precisely, we would need to know the genealogical tree relating all sampled individuals
at each position along the genome.

There is another compelling reason for recording the entire genealogical history.
In many applications, simulations are run for the purpose of comparing summary statistics of the resulting genomes
to those computed from real data,
with the purpose of inferring demographic history or identifying regions under selection.
However, for the same reason that neutral mutations can be laid down afterwards on the sequence of genealogical trees resulting from a simulation,
any summary statistic of genetic variation can be expressed as a statistic of the distribution of genealogical trees along the genome,
with additional noise introduced by mutations.
For instance, mean pairwise sequence diversity, $\theta$,
is most easily thought of as an estimator of $2 \mu t_{TMRCA}$, the mean time to most recent common ancestor
multiplied by twice the mutation rate,
with a relative error that is independent of the actual $t_{TMRCA}$.
As additional measurement noise makes simulation-based inference methods such as ABC or trained classifiers more diffcult,
it could be beneficial to compare the statistics observed in the data
to their expected values conditioned on the genealogical history,
as remaining mutational noise is not informative.
\plr{Say better.  A quick way to demonstrate this point?}

There is an obvious cost to recording all of history from a simulation,
rather than just the current state.
In this paper, we show how to use algorithmic tools and data structures developed for the coalescent simulator \msprime
to efficiently record, and later process, this history.

@@

First we describe the data structures we use for recording genealogical history,
as implemented in \msprime{}.
These derive from those described in \citet{msprime},
but have been modified to remove redundancy.

To clarify terminology,
below a \emph{tree} refers to
a genealogical tree describing how a collection of individuals are related to each other.  
A \emph{tree sequence} contains information sufficient to reconstruct the genealogical
tree relating all samples to each other at any point along the genome.
In the context of a tree sequence,
\emph{nodes} refer to distinct ancestors,
and may be identified with points in the trees of a tree sequence.
Every tip or branching point in a tree is associated with a node.
A tree sequence does not necessarily carry complete genealogical information about every node,
but calling a node a \emph{sample} implies that complete information is present for that ancestor,
i.e., the node is associated with a point in every tree in the tree sequence.
Since each node represents a certain ancestor, it has a
unique ``time``, thought of as her birth time, which determines the height
of any branching points she is associated with.  A given node will be
associated with branching points of all trees across a region if that node
is the most recent common ancestor to the subtending tips across that region.  
This information is stored in the columns of a \textbf{Node Table}:
\begin{align*}
    \text{flags,} \qquad \text{population,} \qquad \text{time} \\ 
\end{align*}
where ``flags'' records other information 
(e.g., a binary mask of `1' indicates the node is a sample).
Importantly, the \textbf{node ID} of a node is given implicitly
by the (zero-based) index of its corresponding row in the Node Table.

Tree sequences are constructed by specifying over which segments of genome
which nodes inherit from which other nodes.  This information is stored by
recording the endpoints of each distinctly inherited ancestral segment,
the parental node, and a list of children nodes who have inherited that segment.
As each such record describes a collection of edges across a swatch of trees in the tree sequence,
we call these \emph{edgesets} 
and store them in the columns of an \textbf{Edgeset Table:}
\begin{align*}
    \text{left,} \qquad \text{right,} \qquad \text{parent,} \qquad \text{children}
\end{align*}

To record information about genetic variants we need to also record each mutation
and which nodes have inherited that mutation.
The tree structure takes care of inheritance -- all we need to do
is to record the highest node in the tree at the mutated site
that inherited that mutation.
As more than one mtuation may occur at a given site,
we separate this information into two tables,
first, the \textbf{Site Table} records for each variant site
\begin{align*}
    \text{position,} \qquad \text{ancestral state}
\end{align*}
Here ``position'' is a (floating point) position along the chromosome,
and ``ancestral state'' is the genotype of the root of the tree at that site.
As for nodes, \textbf{site IDs} are given implicitly
by the (zero-based) index of the rows.
Then, we record in a \textbf{Mutation Table}
\begin{align*}
    \text{site,} \qquad \text{node,} \qquad \text{derived state}
\end{align*}
in which ``site'' is the ID of the site at wihch this mutation occurred,
``node'' is the ID of the highest node that has inherited this mutation,
and ``derived state'' is the genotype at this site of any individuals inheriting this mutation,
unless another mutation occurs.
\plr{Omit migrations?}

@@

The resulting tables encode everything we need to know --
in fact, they record all of history for everyone alive at any time through the simulation.
This is much more than we need to reconstruct the genealogies and sequences
of a smaller sample of indidivuals.
Reducing this larger tree sequence to a smaller one relevant to a given set of ``sample'' nodes
we call \emph{simplification}.
Roughly, this works by tracing ancestry from the samples backwards through the recorded history,
adding node and edgeset records to the output only when coalescent events are reached.
This works exactly as in \msprime, allowing substantial re-use of algorithms;
the main difference being that parental choice, mutations, and recombination locations 
are determined by the input tree sequence
rather than randomly generated.
An example is shown in Figure \ref{fig:method_diagram}.
\plr{(trying to make the point strongly in the figure, that everything can be stored quite concisely)}

\plr{Note that you can begin the forwards-time simulation with the results of a previous (coalescent) one,
which is helpful so you don't need to run for so long (cite Wilkins).}
In this scheme,
at any point in the simulation genealogical history is recorded in a tree sequence.
This has two additional advantages.
First, simplification can be run periodically through the simulation,
taking the set of samples to be the entire currently alive population.
This is important as it keeps memory usage from growing linearly (and quickly) with time.
Second, the simulation can be \emph{begun} with a tree sequence produced by some other method --
for instance, by a coalescent simulation with \msprime.
This allows for incorporation of deep-time history beyond the reach of individual-based simulations.
Since geographic structure from times longer ago than the mixing time of migration across the range
has limited effect on modern genealogies \citep{wilkins_separation}
(other than possibly changing effective population size \citet{durretspatial}),
this may not negatively affect realism.

>>>>

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath, amssymb}
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{listings}
\bibliographystyle{plainnat}
% \usepackage[T1]{fontenc}

@@

However, it is thought that much of that variation is selectively neutral (or nearly so).
By definition, the alleles carried by individuals in a population at neutral sites
do not affect the population process.
For this reason, if one records the entire genealogical history of a population over the course of a simulation,
one can lay down neutral mutations on top of that history afterwards,
without loss of generality.
Precisely, we would need to know the genealogical tree relating all sampled individuals
at each position along the genome.
In this paper, we show how to use algorithmic tools and data structures developed for the coalescent simulator \msprime
to efficiently record, and later process, this history.

@@

First we describe the data structures we use for recording genealogical history,
as implemented in \msprime{}.
These derive from those described in \citet{msprime},
but have been modified to remove redundancy.
The tables below give the example tree sequence of Figure \ref{fig:ex_tree_seq}.

\begin{figure}
    \begin{center}
        \includegraphics[width=0.6\textwidth]{example_tree_seq}
    \end{center}
    \caption{
        A pictorial representation of the ARG
        relating three samples to each other
        over a chromosome of length 1.0.
        Mutations shown in the example tables of the text
        are marked with red stars.
        \label{fig:ex_tree_seq}
    }
\end{figure}

To clarify terminology,
below a \emph{tree} refers to
a genealogical tree describing how a collection of individuals are related to each other.  
A \emph{tree sequence} contains information sufficient to reconstruct the genealogical
tree relating all samples to each other at any point along the genome.
In the context of a tree sequence,
\emph{nodes} refer to distinct ancestors,
and may be identified with points in the trees of a tree sequence.
Every tip or branching point in a tree is associated with a node.
A tree sequence does not necessarily carry complete genealogical information about every node,
but calling a node a \emph{sample} implies that complete information is present for that ancestor,
i.e., the node is associated with a point in every tree in the tree sequence.
Since each node represents a certain ancestor, it has a
unique ``time'', thought of as her birth time, which determines the height
of any branching points she is associated with.  A given node will be
associated with branching points of all trees across a region if that node
is the most recent common ancestor to the subtending tips across that region.  
This information is stored in the columns of a \textbf{Node Table}:
\begin{center}
\begin{tabular}{cccc}
    id   &  is\_sample &  population &  time \\
    0    &  1          & 0           & 0 \\
    1    &  1          & 0           & 0 \\
    2    &  1          & 0           & 0 \\
    3    &  0          & 0           & 0.3 \\
    4    &  0          & 0           & 0.5 \\
    5    &  0          & 0           & 0.6 \\
    6    &  0          & 0           & 0.7 \\
\end{tabular}
\end{center}
where ``flags'' records other information 
(e.g., a binary mask of `1' indicates the node is a sample).
Importantly, the \textbf{node ID} of a node is given implicitly
by the (zero-based) index of its corresponding row in the Node Table.

Tree sequences are constructed by specifying over which segments of genome
which nodes inherit from which other nodes.  This information is stored by
recording the endpoints of each distinctly inherited ancestral segment,
the parental node, and a list of children nodes who have inherited that segment.
As each such record describes a collection of edges across a swatch of trees in the tree sequence,
we call these \emph{edgesets} 
and store them in the columns of an \textbf{Edgeset Table:}
\begin{center}
\begin{tabular}{cccc}
    left & right & parent & children \\
    0.2  &  0.8  &  3     & 0,2 \\
    0.0  &  0.2  &  4     & 1,2 \\
    0.2  &  0.8  &  4     & 1,3 \\
    0.8  &  1.0  &  4     & 1,2 \\
    0.8  &  1.0  &  5     & 0,4 \\
    0.0  &  0.2  &  6     & 0,4  \\
\end{tabular}
\end{center}

To record information about genetic variants we need to also record each mutation
and which nodes have inherited that mutation.
The tree structure takes care of inheritance -- all we need to do
is to record the highest node in the tree at the mutated site
that inherited that mutation.
As more than one mtuation may occur at a given site,
we separate this information into two tables,
first, the \textbf{Site Table} records for each variant site
\begin{center}
\begin{tabular}{ccc}
    id  &   position  & ancestral\_state \\
    0   &  0.1        & 0 \\
    1   &  0.5        & 0 \\
\end{tabular}
\end{center}
Here ``position'' is a (floating point) position along the chromosome,
and ``ancestral state'' is the genotype of the root of the tree at that site.
As for nodes, \textbf{site IDs} are given implicitly
by the (zero-based) index of the rows.
Then, we record in a \textbf{Mutation Table}
\begin{center}
\begin{tabular}{ccc}
    site  & node  & derived\_state \\
    0	  & 4	  & 1 \\
    1	  & 3	  & 1 \\
    1	  & 2	  & 0 \\
\end{tabular}
\end{center}
in which ``site'' is the ID of the site at wihch this mutation occurred,
``node'' is the ID of the highest node that has inherited this mutation,
and ``derived state'' is the genotype at this site of any individuals inheriting this mutation,
unless another mutation occurs.
% \plr{Omit migrations?}

@@

The resulting tables encode everything we need to know --
in fact, they record all of history for everyone alive at any time through the simulation.
This is much more than we need to reconstruct the genealogies and sequences
of a smaller sample of indidivuals.
Reducing this larger tree sequence to a smaller one relevant to a given set of ``sample'' nodes
we call \emph{simplification}.
Roughly, this works by tracing ancestry from the samples backwards through the recorded history,
adding node and edgeset records to the output only when coalescent events are reached.
This works exactly as in \msprime, allowing substantial re-use of algorithms;
the main difference being that parental choice, mutations, and recombination locations 
are determined by the input tree sequence
rather than randomly generated.
An example is shown in Figure \ref{fig:method_diagram}.

In this scheme,
at any point in the simulation genealogical history is recorded in a tree sequence.
This has two additional advantages.
First, simplification can be run periodically through the simulation,
taking the set of samples to be the entire currently alive population.
This is important as it keeps memory usage from growing linearly (and quickly) with time.
Second, the simulation can be \emph{begun} with a tree sequence produced by some other method --
for instance, by a coalescent simulation with \msprime.
This allows for incorporation of deep-time history beyond the reach of individual-based simulations.
Since geographic structure from times longer ago than the mixing time of migration across the range
has limited effect on modern genealogies \citep{wilkins_separation}
(other than possibly changing effective population size \citet{durretspatial}),
this may not negatively affect realism.
